Mental Rotation as Bayesian Quadrature

### 2013-05-31 man/nips-2013.tex
First submission version

\begin{abstract}
  Given a computationl resource--for example, the ability to visualize
  an object rotating--how do you best make use of it? We explored how
  mental simulation should used in the classic psychological task of
  determining if two images depict the same object in different
  orientations. We compared three models on this mental rotation task,
  and found that a model based on an optimal experiment design for
  Bayesian quadrature is objectively more accurate and qualitatively
  more consistent with classic behavioral data than the other models
  we evaluated. We suggest that rational models which adaptively
  exploit available resources are promising in their ability to
  characterize metacognitive processes like mental simulation.
\end{abstract}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/shepard-rotation.png}
  \caption{\textbf{Classic mental rotation task}. Participants in
    \cite{Shepard1971} saw stimuli such as those on the left, and
    judged whether each pair of shapes was the same shape in two
    different orientations (``same'' pairs), or two different shapes
    (``different'' pairs). \textbf{A} and \textbf{B} show ``same''
    pairs, while \textbf{C} shows a ``different'' pair. The plots on
    the right indicate mean response times on the mental rotation
    task, exhibiting a strong linear relationship with increasing
    variance as a function of true rotation.}
  \label{fig:mental-rotation}
\end{figure}

@@

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Response error rates}: How accurate was the model at
  choosing the correct hypothesis? This was defined to be the mean
  error ($\ME{}$), or fraction of times the model picks the incorrect
  hypothesis.
\item \textit{Rotations}: For those ``same'' pairs which the model
  judged correct, how correlated were the model's rotations with the
  true angles of rotation?  We quantified this using the Pearson's
  correlation coefficient $\rho$ for the true rotation ($R$) versus
  the extent of the model's rotations ($\vert \mathbf{R}\vert$). The
  top row in Figure \ref{fig:rotations} shows individual points
  corresponding to true rotations vs. rotations by the models for all
  stimuli, including those judged incorrectly. The bottom row shows
  mean model rotations, across only those stimuli which were judged
  correctly. We structured the analysis in this way to better compare
  to the results of \cite{Shepard1971}, which also excluded
  incorrectly judged stimuli.
\item \textit{Estimates of $Z$}: How accurate was the model's estimate
  of $Z$? We defined this quantity as the mean squared error
  ($\MSE{}$) between the model's estimate of $Z$ and the ``gold
  standard'' value, where the error has scale such that $\MSE{}=0$
  indicates no error and $\MSE{}=1$ indicates maximum error. Figure
  \ref{fig:accuracy} shows plots of the true (``gold standard'') value
  of $Z$ vs. the model's estimate, for each model.
\end{itemize}

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). Closer inspection reveals that much of this comes
from ``different'' pairs ($\NaiveMEdiff{}$), where the asymmetric
linear interpolation may give an overestimate of $Z$ (e.g. Figure
\ref{fig:li}).  This intuition is confirmed by the model's accuracy in
estimating $Z$, which was $\NaiveMSE{}$. As shown in the left panel of
Figure \ref{fig:accuracy}, it overestimated $Z$ for nearly all stimuli
pairs.

The correlation between the \naive{} model's average rotation and the
true angle of rotation was $\Naivecorr{}$ (Figure \ref{fig:rotations},
bottom left). More complex patterns are revealed by examining the
shape of the raw data in Figure \ref{fig:rotations}, top left: the
\naive{} model corresponds extremeley well to the true angle of
rotation for approximately $R<\frac{\pi}{2}$. This is unsurprising,
because the closer the true angle is to zero, the less the model has
to rotate, and the less likely it will get stuck on local
maxima. Thus, it is more likely to locate the global maximum, which
corresponds to the true angle of rotation. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The average rotations computed by the nonparametric model were
strongly correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, bottom right). Because the nonparametric model
has the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g. Figure \ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident that its
estimate of $Z$ is accurate, it does not get stuck as easily on local
optima.  Indeed, its active attempts to obtain an accurate estimate of
$Z$ were largely succeessful, with $\BQMSE{}$.

@@

In this paper, we asked: how do people use mental simulation?  We
investigated the specific case of mental rotation, taking a rational
analysis approach to characterizing optimal strategies for performing
mental rotation. We found that an adaptive, nonparametric model
performs best at the classic rotation task \cite{Shepard1971}, as it
is able to actively monitor the confidence of its estimate and
intelligently choose the direction of rotation. Two simpler models
based on heuristics performed much worse: they did not maintain the
linear relationship with the true angle of rotation due to lack of
robustness against local maxima or incorrect rotation direction. The
nonparametric model qualitatively approximates human behavioral
judgments, both in linearity \cite{Shepard1971} and variability
\cite{Just1976}. Future work will collect empirical data from
participants to perform a more thorough, quantitative analysis.  We
conclude that from this initial survey, models of mental rotation
which take an ``active'', directed approach seem well-suited to
explaining human behavior in such metacognitive domains as mental
simulation.

@@

% \renewcommand\bibsection{\subsubsection*{\refname}}
\renewcommand\refname{\normalsize{References}}
\bibliographystyle{ieeetr}
\bibliography{references}

>>>>

\begin{abstract}
  Given a computational resource--for example, the ability to
  visualize an object rotating--how do you best make use of it? We
  explored how mental simulation should used in the classic
  psychological task of determining if two images depict the same
  object in different orientations. We compared three models on this
  mental rotation task, and found that a model based on an optimal
  experiment design for Bayesian quadrature is objectively more
  accurate and qualitatively more consistent with classic behavioral
  data than the other models we evaluated. We suggest that rational
  models which adaptively exploit available resources are promising in
  their ability to characterize metacognitive processes like mental
  simulation.
\end{abstract}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.7\textwidth]{../figures/shepard-rotation.png}
  \caption{\textbf{Classic mental rotation task}. Participants in
    \cite{Shepard1971} saw stimuli such as those on the left, and
    judged whether each pair of shapes was the same shape in two
    different orientations (``same'' pairs), or two different shapes
    (``different'' pairs). \textbf{A} and \textbf{B} show ``same''
    pairs, while \textbf{C} shows a ``different'' pair. The plots on
    the right indicate mean response times on the mental rotation
    task, exhibiting a strong linear relationship with increasing
    variance as a function of true rotation.}
  \label{fig:mental-rotation}
\end{figure}

@@

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Response error rates}: How accurate was the model at
  choosing the correct hypothesis? This was defined as the mean error
  ($\ME{}$), or fraction of times the model chose incorrectly.
\item \textit{Rotations}: For those ``same'' pairs which the model
  judged correct, how correlated were the model's rotations with the
  true angles of rotation?  We quantified this using the Pearson's
  correlation coefficient $\rho$ for the true rotation ($R$) versus
  the extent of the model's rotations ($\vert \mathbf{R}\vert$). The
  top row in Figure \ref{fig:rotations} shows individual points
  corresponding to true rotations vs. rotations by the models for all
  stimuli, including those judged incorrectly. The bottom row shows
  mean model rotations, across only those stimuli which were judged
  correctly. We structured the analysis in this way to better compare
  to the results of \cite{Shepard1971}, which also excluded
  incorrectly judged stimuli.
\item \textit{Estimates of $Z$}: How accurate was the model's estimate
  of $Z$? We defined this quantity as the mean squared error
  ($\MSE{}$) between the model's estimate of $Z$ and the ``gold
  standard'' value, where the error has scale such that $\MSE{}=0$
  indicates no error and $\MSE{}=1$ indicates maximum error. Figure
  \ref{fig:accuracy} shows plots of the true (``gold standard'') value
  of $Z$ vs. the model's estimate, for each model.
\end{itemize}

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). Closer inspection reveals that much of this comes
from ``different'' pairs ($\NaiveMEdiff{}$), where the asymmetric
linear interpolation may give an overestimate of $Z$ (e.g. Figure
\ref{fig:li}).  This intuition is confirmed by the model's accuracy in
estimating $Z$, which was $\NaiveMSE{}$. As shown in the left panel of
Figure \ref{fig:accuracy}, it overestimated $Z$ for nearly all
stimuli.

The correlation between the \naive{} model's average rotation and the
true angle of rotation was $\Naivecorr{}$ (Figure \ref{fig:rotations},
bottom left). More complex patterns are revealed by examining the
shape of the raw data in Figure \ref{fig:rotations}, top left: the
\naive{} model corresponds extremely well to the true angle of
rotation for approximately $R<\frac{\pi}{2}$. This is unsurprising,
because the closer the true angle is to zero, the less the model has
to rotate, and the less likely it will get stuck on local
maxima. Thus, it is more likely to locate the global maximum, which
corresponds to the true angle of rotation. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The average rotations computed by the nonparametric model were
strongly correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, bottom right). Because the nonparametric model
has the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g. Figure \ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident that its
estimate of $Z$ is accurate, it does not get stuck as easily on local
optima.  Indeed, it was largely successful, with $\BQMSE{}$.

@@

In this paper, we asked: how do people use mental simulation?  We
investigated the specific case of mental rotation, taking a rational
analysis approach to characterizing optimal strategies for performing
mental rotation. We found that an adaptive, nonparametric model
performs best at the classic rotation task \cite{Shepard1971}, as it
is able to actively monitor the confidence of its estimate and
intelligently choose the direction of rotation. Two simpler models
based on heuristics performed much worse: they did not maintain the
linear relationship with the true angle of rotation due to lack of
robustness against local maxima or incorrect rotation direction. The
nonparametric model qualitatively approximates human behavioral
judgments, both in linearity \cite{Shepard1971} and variability
\cite{Just1976}. Future work will collect empirical data from
participants to perform a more thorough, quantitative analysis.  We
conclude that from this initial survey, models which take an
``active'', directed approach seem well-suited to explaining mental
simulation.

@@

% \renewcommand\bibsection{\subsubsection*{\refname}}
\renewcommand\refname{\normalsize{References}}
\bibliographystyle{ieeetr}
{\small \bibliography{references}}
